### 2020-04-04 reports/thesis/chapters/4_variancePrediction.tex
Flowchart is png
Corrections

It was found that it was hard to fit a compound Poisson distribution onto the grey values. Results did show that the variance of the grey value has a linear relationship with the mean grey value. As a result, it would be possible to predict the variance of the grey value, given the grey value. This opens up new ways to predict the uncertainty of each pixel in an x-ray projection.

@@

The sample variance-mean data were obtained from the replicate projections. Let $x_{i,j}^{(r)}$ be the grey value of the pixel at $(i,j)$ from the $r$th replicate projection for $r=1,2,3,\dotsc,m$. The sample mean and sample variance grey value for the pixel at $(i,j)$ are
\begin{equation}
    \overline{x}_{i,j}=\frac{1}{m}\sum_{r=1}^m x_{i,j}^{(r)}
\end{equation}
and
\begin{equation}
    y_{i,j} =
    \frac{1}{m-1}
    \sum_{r=1}^m
        \left(
            x_{i,j}^{(r)} - \overline{x}_{i,j}
        \right)^2
\end{equation}
respectively. Only pixels in the region of interest (ROI) were considered here, that is, pixels which represent the test sample. The ROI was created by manually segmenting the test sample from the projection.

The aim is to model and predict the grey value variance of a pixel given its grey value by using a model fitted onto the sample variance-mean data. No spatial information was used, thus, the variance-mean data is denoted as $(x_1,y_1),(x_2,y_2),\dotsc,(x_N,y_N)$ from now on, where $N$ is the area of the ROI or the size of the dataset. Let $Y(x)$ be a random variable and the sample variance given a grey value mean $x$. It was assumed that the standard error from estimating the mean was negligible so that the uncertainty is captured by the random variable $Y$. It was assumed that for a given pixel, the grey values are Normal and i.i.d. Let $\sigma^2(x)$ be the variance given a grey value $x$, then it can be shown that
\begin{equation}
\dfrac{(m-1)Y(x)}{\sigma^2(x)}\sim\chi^2_{m-1}
\end{equation}
which results in
\begin{equation}
Y(x)\sim\gammaDist\left(\alpha,\dfrac{\alpha}{\sigma^2(x)}\right)
\end{equation}
where $\alpha=(m-1)/2$ is the shape parameter. The expectation and variance are
\begin{equation}
\expectation\left[Y(x)\right]=\sigma^2(x)
\end{equation}
and
\begin{equation}
\variance\left[Y(x)\right]=\left(\dfrac{\sigma^2(x)}{\alpha}\right)^2
\end{equation}
respectively.

This framework allows the use of generalised linear models (GLM) \citep{nelder1972generalized,nelder1972generalized_2, mccullagh1984generalized}. In the gamma distribution case, a GLM can be used to model
\begin{equation}
Y(x)\sim\gammaDist\left(\alpha,\dfrac{\alpha}{g^{-1}(\eta(x))}\right)
\end{equation}
where $g(y)$ is the link function and $\eta(x)$ is a linear function called the systematic component. It should be noted that
\begin{equation}
  \expectation\left[Y(x)\right]=g^{-1}(\eta(x))
\end{equation}
which shows how the link function and systematic component work together. Examples of link functions are the identity link
\begin{equation}
g(y)=g^{-1}(y)=y
\end{equation}
and, for the gamma distribution case, the canonical link
\begin{equation}
g(y)=g^{-1}(y)=1/y \ .
\end{equation}
There is also the log link $g(y)=\ln y$ but $y$ must be a unitless quantity, for example Poisson counts, for it to be comparable. Because $y$ is in units of $\SI{}{\adu^2}$, the log link was not used in this analysis.

An example of a systematic component are polynomial features $\eta(x)=\sum_{r=0}^{p}\beta_r x^{r}$ so that when used with the identity link, for example, $\expectation\left[Y(x)\right] = \sum_{r=0}^{p}\beta_r x^{r}$. Iterative reweighted least squares \citep{friedman2001elements} can be used to estimate the parameters $\beta_0, \beta_1, \dotsc, \beta_{p}$ given data for the model to fit onto.

@@

Figures \ref{fig:meanVar_varMeanExample_AbsNoFilter} and \ref{fig:meanVar_varMeanExample_AbsFilter} shows the GLM fits for the datasets \texttt{AbsNoFilter} and \texttt{AbsFilter} respectively. The prediction intervals were obtained using the distribution $\widehat{Y}(x)\sim\gammaDist\left(\alpha, \dfrac{\alpha}{g^{-1}\left(\widehat{\eta}(x)\right)}\right)$. The fits all looked reasonable except for \texttt{AbsNoFilter} at \ang{120} where the fit did not capture the inflection. The forward stepwise selection may not picked this up because there were a lot of low grey values, causing a leverage towards these low grey values. As a result, the inflection did not stood out to the method.

@@

Instead of assessing the model fit using the AIC and BIC, the performance on predicting the variance given a grey value was assessed using cross-validation \citep{friedman2001elements}. Cross-validation assesses the model to make predictions on data it has not seen before. This was done by randomly splitting the $N$ data points into two disjoint sets, the training set and the test set $\tau$. It was chosen that the training and test set are of the same size $N_\tau$. The model was fitted onto the training set. Afterwards, the fitted model predicts the variances given the grey values in the test set which were then compared to the actual variances.

@@

\begin{figure}
  \centering
  \centerline{
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsNoFilterDeg30_vsgreyvalue.eps}
      \caption{\texttt{AbsNoFilter} at \ang{30}}
  \end{subfigure}
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsNoFilterDeg120_vsgreyvalue.eps}
      \caption{\texttt{AbsFilter} at \ang{120}}
  \end{subfigure}
  }
  \centerline{
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsFilterDeg30_vsgreyvalue.eps}
      \caption{\texttt{AbsFilter} at \ang{30}}
  \end{subfigure}
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsFilterDeg120_vsgreyvalue.eps}
      \caption{\texttt{AbsFilter} at \ang{120}}
  \end{subfigure}
  }
  \caption{Log frequency density histogram of the residuals given the mean grey value. The residuals are from fitting a gamma GLM of the form ${y}(x)=\beta_0+\beta_1 x$ onto the sample variance-mean data obtained from the replicate projections. The dashed red line shows the 68\% prediction residual interval. The colour scales are in units of $\log\SI{}{\adu^{-3}}$.}
  \label{fig:meanVar_residualGreyvalue}
\end{figure}

@@

It was unusual to see an infection in the sample variance-mean data in \texttt{AbsNoFilter} at \ang{120} which was not captured by the GLM. A good explanation was not found but perhaps the absence of an x-ray filter may have contributed to this.

More flexible models such as non-parametric and machine learning methods may be used, however, they are slow and are unnecessary in a low dimensional problem with a large number of data points.

>>>>

In the previous chapter, the grey values were modelled using a compound Poisson distribution to reflect the science of x-rays. This model has a linear relationship between the variance and the expectation. As a result, it would be possible to predict the variance of the grey value, given the grey value. This opens up new ways to predict the uncertainty of each pixel in an x-ray projection.

@@

The sample variance-mean data were obtained from the replicate projections. Let $x_{i,j}$ be the grey value of the $i$th pixel, ignoring any spatial information, from the $j$th replicate projection for $i=1,2,3,\ldots,N$ and $j=1,2,3,\ldots,m$. The sample mean and sample variance grey value for the $i$th pixel are
\begin{equation}
    \overline{x}_{i}=\frac{1}{m}\sum_{j=1}^m x_{i,j}
\end{equation}
and
\begin{equation}
    y_{i} =
    \frac{1}{m-1}
    \sum_{j=1}^m
        \left(
            x_{i,j} - \overline{x}_{i}
        \right)^2
\end{equation}
respectively. The symbol $y$ should not be confused with the latent varible in the previous chapter. Only pixels in the region of interest (ROI) were considered here, that is, pixels which represent the test sample. The ROI was created by manually segmenting the test sample from the projection.

The aim is to model and predict the grey value variance of a pixel given its grey value by using a model fitted onto the sample variance-mean data. The variance-mean data is denoted as $(x_1,y_1),(x_2,y_2),\ldots,(x_N,y_N)$ where $N$ is the area of the ROI or the size of the dataset. Let $Y(x)$ be a random variable and the sample variance given a grey value mean $x$. It was assumed that the standard error from estimating the mean was negligible so that the uncertainty is captured by the random variable $Y$. It was assumed that for a given pixel, the grey values are Normal and i.i.d. Let $\sigma^2(x)$ be the variance given a grey value $x$, then it can be shown that
\begin{equation}
\dfrac{(m-1)Y(x)}{\sigma^2(x)}\sim\chi^2_{m-1}
\end{equation}
which results in
\begin{equation}
Y(x)\sim\gammaDist\left(\alpha,\dfrac{\alpha}{\sigma^2(x)}\right)
\end{equation}
where $\alpha=(m-1)/2$ is the shape parameter. The expectation and variance are
\begin{equation}
\expectation\left[Y(x)\right]=\sigma^2(x)
\end{equation}
and
\begin{equation}
\variance\left[Y(x)\right]=\left(\dfrac{\sigma^2(x)}{\alpha}\right)^2
\end{equation}
respectively.

This framework allows the use of generalised linear models (GLM) \citep{nelder1972generalized,nelder1972generalized_2, mccullagh1984generalized}. In the gamma distribution case, a GLM can be used to model
\begin{equation}
Y(x)\sim\gammaDist\left(\alpha,\dfrac{\alpha}{g^{-1}(\eta(x))}\right)
\end{equation}
where $g(y)$ is the link function and $\eta(x)$ is a linear function called the systematic component. It should be noted that
\begin{equation}
  \expectation\left[Y(x)\right]=g^{-1}(\eta(x))
\end{equation}
which shows how the link function and systematic component work together. Examples of link functions are the identity link
\begin{equation}
g(y)=g^{-1}(y)=y
\end{equation}
and, for the gamma distribution case, the canonical link
\begin{equation}
g(y)=g^{-1}(y)=1/y \ .
\end{equation}

An example of a systematic component are polynomial features $\eta(x)=\sum_{r=0}^{p}\beta_r x^{r}$ so that when used with the identity link, for example, $\expectation\left[Y(x)\right] = \sum_{r=0}^{p}\beta_r x^{r}$. Iterative reweighted least squares \citep{friedman2001elements} can be used to estimate the parameters $\beta_0, \beta_1, \ldots, \beta_{p}$ given data for the model to fit onto.

@@

Figures \ref{fig:meanVar_varMeanExample_AbsNoFilter} and \ref{fig:meanVar_varMeanExample_AbsFilter} shows the GLM fits for the datasets \texttt{AbsNoFilter} and \texttt{AbsFilter} respectively. The prediction intervals were obtained using the distribution $\widehat{Y}(x)\sim\gammaDist\left(\alpha, \dfrac{\alpha}{g^{-1}\left(\widehat{\eta}(x)\right)}\right)$. The fits all looked reasonable except for \texttt{AbsNoFilter} at \ang{120} where the fit did not capture the inflection. The forward stepwise selection may not picked this up because there were a lot of low grey values, causing a leverage towards these low grey values. As a result, the inflection did not stand out to the method.

@@

Instead of assessing the model fit using the AIC and BIC, the performance on predicting the variance given a grey value was assessed using cross-validation \citep{allen1974relationship, stone1974cross, stone1977asymptotic, friedman2001elements}. Cross-validation assesses the model to make predictions on data it has not seen before. This was done by randomly splitting the $N$ data points into two disjoint sets, the training set and the test set $\tau$. The model was fitted onto the training set. Afterwards, the fitted model predicts the variances given the grey values in the test set which were then compared to the actual variances.

It was chosen that the training and test set are of the same size, a 50:50 spilt. The sizes of the sets can be different, for example, a 75:25 spilt. A 50:50 spilt was chosen because a large training set is an overkill for a high data and low dimensional problem. Here, a model with a few parameters is fitted onto millions of pixels.

@@

\begin{figure}
  \centering
  \centerline{
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsNoFilterDeg30_vsgreyvalue.eps}
      \caption{\texttt{AbsNoFilter} at \ang{30}}
  \end{subfigure}
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsNoFilterDeg120_vsgreyvalue.eps}
      \caption{\texttt{AbsNoFilter} at \ang{120}}
  \end{subfigure}
  }
  \centerline{
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsFilterDeg30_vsgreyvalue.eps}
      \caption{\texttt{AbsFilter} at \ang{30}}
  \end{subfigure}
  \begin{subfigure}[b]{\subSize}
      \includegraphics[width=\textwidth]{../figures/varmean/varMeanResidualAbsFilterDeg120_vsgreyvalue.eps}
      \caption{\texttt{AbsFilter} at \ang{120}}
  \end{subfigure}
  }
  \caption{Log frequency density histogram of the residuals given the mean grey value. The residuals are from fitting a gamma GLM of the form ${y}(x)=\beta_0+\beta_1 x$ onto the sample variance-mean data obtained from the replicate projections. The dashed red line shows the 68\% prediction residual interval. The colour scales are in units of $\log\SI{}{\adu^{-3}}$.}
  \label{fig:meanVar_residualGreyvalue}
\end{figure}

@@

It was unusual to see an infection in the sample variance-mean data in \texttt{AbsNoFilter} at \ang{120} which was not captured by the GLM. A good explanation was not found but perhaps the absence of an x-ray filter may have contributed to this. As a result, the \texttt{AbsNoFilter} dataset was discarded in favour of the \texttt{AbsFilter} dataset from the next chapter.

There exist model selection methods such as lasso and elastic net \citep{tibshirani1996regression, zou2005regularization, friedman2010regularization} but they are catered for high dimensional problems involving hundreds of parameters. In this problem, only a few parameters were needed to find a good model. More flexible models such as non-parametric and machine learning methods may be used, however, they are slow and are unnecessary in a low dimensional problem with a large number of data points.
